---
title: "wild/domesticated sunflower RNAseq: WGCNA"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
params:
  rmd: ""
output:
  html_document:
    dev: png
    code_folding: hide
    self_contained: yes
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vroom)
library(tidyverse)
library(WGCNA)
```

# Data input and cleaning

```{r import_data}
normLogCPM <- vroom("../deliverables/norm_log_cpm.tsv") %>% 
  dplyr::mutate(gene=gsub("gene:","",gene)) %>%
  column_to_rownames(var="gene")

# reformat to wide df (rows=samples,columns=genes)
datExpr0 <- t(normLogCPM)
rm(normLogCPM)
```

```{r data_cleaning}
gsg <- goodSamplesGenes(datExpr0, verbose = 3)

# If the last statement returns TRUE, all genes have passed the cuts. If not, we remove the offending genes and samples from the data:
if (!gsg$allOK){
  # Optionally, print the gene and sample names that were removed:
  if (sum(!gsg$goodGenes)>0)
     printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
  if (sum(!gsg$goodSamples)>0)
     printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
  # Remove the offending genes and samples from the data:
  datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}

```

```{r cluster_samples}
sampleTree <- hclust(dist(datExpr0), method = "average")
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small. sizeGrWindow(12,9)
#pdf(file = "Plots/sampleClustering.pdf", width = 12, height = 9);
par(cex = 0.6)
par(mar = c(0,4,2,0))
plot(sampleTree,
     main = "Sample clustering to detect outliers",
     sub="", xlab="", cex.lab = 1.5,cex.axis = 1.5, cex.main = 2)
```
It looks like there aren't any outliers, so we will continue with the dataframe as-is.
```{r remove_outliers}
datExpr <- datExpr0
rm(datExpr0)
```

We now read in the trait data and match the samples for which they were measured to the expression samples
```{r load_binarized_trait_data}
# import metadata
allTraits <- vroom("units.tsv") %>% 
  # remove unneccessary columns
  dplyr::select(sample) %>%
  dplyr::mutate(
    # add binary columns for accession
    arikara=ifelse(grepl("arikara",sample),1,0),
    hidatsa=ifelse(grepl("hidatsa",sample),1,0),
    hopi=ifelse(grepl("hopi",sample),1,0),
    mandan=ifelse(grepl("mandan",sample),1,0),
    seneca=ifelse(grepl("seneca",sample),1,0),
    nebraska=ifelse(grepl("nebraska",sample),1,0),
    arkansas=ifelse(grepl("arkansas",sample),1,0),
    tennessee=ifelse(grepl("tennessee",sample),1,0),
    kentucky=ifelse(grepl("kentucky",sample),1,0),
    missouri=ifelse(grepl("missouri",sample),1,0),
    # add binary columns for tissue
    endosperm=ifelse(grepl("endosperm",sample),1,0),
    seedcoat=ifelse(grepl("seedcoat",sample),1,0),
    earlyBudding=ifelse(grepl("earlyBudding",sample),1,0),
    lateBudding=ifelse(grepl("lateBudding",sample),1,0),
    leaf=ifelse(grepl("leaf",sample),1,0),
    meristem=ifelse(grepl("meristem",sample),1,0),
    node=ifelse(grepl("node",sample),1,0),
    root=ifelse(grepl("root",sample),1,0),
    shoot=ifelse(grepl("shoot",sample),1,0),
    # add binary column for domestication
    domestication=case_when(
      arikara==1 | hidatsa==1 | hopi==1 | mandan==1 | seneca==1 ~ 1,
      nebraska==1 | arkansas==1 | tennessee==1 | kentucky==1 | missouri==1 ~ 0
    )
  )

# get sample_names in order of logCPM matrix
sample_names <- rownames(datExpr)
# match 
traitRows <- match(sample_names, allTraits$sample)
datTraits <- allTraits %>% dplyr::select(-c("sample")) %>% .[traitRows,]

```

We now have the expression data in the variable `datExpr`, and the corresponding clinical traits in the variable `datTraits`. Before we continue with network construction and module detection, we visualize how the clinical traits relate to the sample dendrogram.
```{r relate_traits_to_dend,fig.width=12,fig.height=12}
# Re-cluster samples
sampleTree2 <- hclust(dist(datExpr), method = "average")
# Convert traits to a color representation: white means low, red means high, grey means missing entry 
traitColors <- numbers2colors(datTraits, signed = FALSE)
# Plot the sample dendrogram and the colors underneath.

sizeGrWindow(12,9)
par(cex = 0.6)
par(mar = c(0,4,2,0))
plotDendroAndColors(sampleTree2,
                    traitColors,
                    groupLabels = names(datTraits),
                    main = "Sample dendrogram and trait heatmap")
```

# One-Step Network Construction and Module Detection

Constructing a weighted gene network entails the choice of the soft thresholding power β to which co-expression similarity is raised to calculate adjacency [1]. The authors of [1] have proposed to choose the soft thresholding power based on the criterion of approximate scale-free topology. We refer the reader to that work for more details; here we illustrate the use of the function pickSoftThreshold that performs the analysis of network topology and aids the user in choosing a proper soft-thresholding power. The user chooses a set of candidate powers (the function provides suitable default values), and the function returns a set of network indices that should be inspected, for example as follows:

```{r choose_soft_thresholding_power}
# Choose a set of soft-thresholding powers
powers <- c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft <- pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
```

```{r plot_thresholding_power}
# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1,2))
cex1 = 0.9

# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"))
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2], labels=powers,cex=cex1,col="red")

# this line corresponds to using an R^2 cut-off of h
abline(h=0.80,col="red")
# Mean connectivity as a function of the soft-thresholding power 
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```

```{r call_modules}
net <- blockwiseModules(datExpr,
                        power = 7,
                        TOMType = "signed",
                        minModuleSize = 30,
                        reassignThreshold = 0,
                        mergeCutHeight = 0.25,
                        numericLabels = TRUE,
                        pamRespectsDendro = FALSE,
                        saveTOMs = TRUE,
                        saveTOMFileBase = "domesticationTOM",
                        verbose = 3)
```

```{r show_module_contents}
table(net$colors)
```


```{r plot_dend_w_modules}
# open a graphics window
sizeGrWindow(12, 9)
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                  "Module colors",
dendroLabels = FALSE, hang = 0.03, addGuide = TRUE, guideHang = 0.05)
```

```{r save_modules_eigengenes}
moduleLabels <- net$colors
moduleColors <- labels2colors(net$colors)
MEs <- net$MEs
geneTree <- net$dendrograms[[1]]
```


# Relating Modules to Trait Data

```{r prepare_data_for_trait_corr}
# Define numbers of genes and samples
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)

# Recalculate MEs with color labels
MEs0 = moduleEigengenes(datExpr, moduleColors)$eigengenes
MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples)
```

```{r module_trait_heatmap,fig.height=20,fig.width=8}
# Will display correlations and their p-values
textMatrix <-  paste(signif(moduleTraitCor, 2), "\n(",
                     signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) <- dim(moduleTraitCor)

par(mar = c(6, 8.5, 3, 3))
# Display the correlation values within a heatmap plot 
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(datTraits),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))

```

```{r ME_corr_to_tibble}
# get labeledHeatmap data for trait in tidy format
ME.tibble <- dplyr::left_join(reshape2::melt(signif(moduleTraitCor, 2), variable.name="ME", value.name="correlation"),
                                     reshape2::melt(signif(moduleTraitPvalue, 2), value.name="pvalue"),
                                     by=c("Var1","Var2")) %>%
  dplyr::rename(ME=Var1,trait=Var2) 

# get top 5 hits for each trait
ME.tophits <- ME.tibble %>% 
  group_by(trait) %>% 
  arrange(correlation) %>% 
  top_n(n=5, wt= abs(correlation)) %>% 
  arrange(trait)
saveRDS(ME.tophits,"ME.tophits.RDS")
```

```{r }
# Define variable weight containing the weight column of datTrait
domestication = as.data.frame(datTraits$domestication)
names(domestication) = "domestication"

# names (colors) of the modules
modNames <- substring(names(MEs), 3)
geneModuleMembership <- as.data.frame(cor(datExpr, MEs, use = "p"))
MMPvalue <- as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples))

names(geneModuleMembership) <- paste("MM", modNames, sep="")
names(MMPvalue) = paste("p.MM", modNames, sep="")

geneTraitSignificance <- as.data.frame(cor(datExpr, domestication, use = "p"))
GSPvalue <- as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));
names(geneTraitSignificance) <- paste("GS.", names(domestication), sep="")
names(GSPvalue) <- paste("p.GS.", names(domestication), sep="")
```

## Intramodular analysis: identifying genes with high Gene Significance and Module Membership

Strong correlations indicate that genes highly significantly associated with a trait are often also the most important (central) elements of modules associated with the trait.

```{r}

res = NULL
for (module in ME.tophits %>% dplyr::mutate(ME=gsub("ME","",ME)) %>% dplyr::filter(trait=="domestication") %>% dplyr::pull(ME)){

  column = match(module, modNames)
  moduleGenes = moduleColors==module

  sizeGrWindow(7, 7)
  par(mfrow = c(1,1))
  
  verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                     abs(geneTraitSignificance[moduleGenes, 1]),
                     xlab = paste("Module Membership in", module, "module"),
                     ylab = "Gene significance for domestication",
                     main = paste("Module membership vs. gene significance\n"),
                     cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)
  
  res <- rbind(res,
               cbind(module, names(as.data.frame(datExpr))[moduleColors==module])
               )
}

# unique(as.data.frame(res)$module)

```

# Save Module Membership Results
```{r}
annot <- vroom("HanXRQr2.0-SUNRISE-2.1.Blast2GO-20181213.table.tsv")
probes2annot <- match(names(as.data.frame(datExpr)), annot$SeqName)

# Create the starting data frame
geneInfo0 <- data.frame(geneID = names(as.data.frame(datExpr)),
                       geneDescription = annot$Description[probes2annot],
                       moduleColor = moduleColors,
                       geneTraitSignificance,
                       GSPvalue)

# Order modules by their significance for weight
modOrder = order(-abs(cor(MEs, domestication, use = "p")))

# Add module membership information in the chosen order 
for (mod in 1:ncol(geneModuleMembership)){
  oldNames = names(geneInfo0)
  geneInfo0 = data.frame(geneInfo0, geneModuleMembership[, modOrder[mod]],
  MMPvalue[, modOrder[mod]]);
  names(geneInfo0) = c(oldNames, paste("MM.", modNames[modOrder[mod]], sep=""),
  paste("p.MM.", modNames[modOrder[mod]], sep=""))
}
# Order the genes in the geneInfo variable first by module color, then by geneTraitSignificance
geneOrder <- order(geneInfo0$moduleColor, -abs(geneInfo0$GS.domestication)); geneInfo = geneInfo0[geneOrder, ]
geneInfo <- geneInfo0[geneOrder, ]
rm(geneInfo0)

write_tsv(geneInfo, file= "wgcna.geneInfo.tsv")
```

```{r}
# save entire environment load("wgcna.RData")
save.image(file="wgcna.RData")
# save RDS for objects used in subsequent chunks
saveRDS(datExpr,"datExpr.RDS")
saveRDS(moduleColors,"moduleColors.RDS")
saveRDS(ME.tophits,"ME.tophits.RDS")
saveRDS(net,"net.RDS")
# clear all objects in global environment to free space
```

# Intramodular Connectivity

```{r}
library(vroom)
library(tidyverse)
library(WGCNA)
```

```{r readRDS_datExpr_moduleColors}
datExpr <- readRDS("datExpr.RDS")
moduleColors <- readRDS("moduleColors.RDS")
ME.tophits <- readRDS("ME.tophits.RDS")
net <- readRDS("net.RDS")
geneInfo <- vroom::vroom("wgcna.geneInfo.tsv")

```


The function `intramodularConnectivity` computes the whole network connectivity `kTotal`, the within module connectivity `kWithin`, `kOut=kTotal-kWithin`, and `kDiff=kIn-kOut=2*kIN-kTotal`
```{r intramodular_connectivity}
ADJ1=abs(cor(datExpr,use="p"))^6
#Alldegrees1=intramodularConnectivity(ADJ1, moduleColors)
#head(Alldegrees1)

ME.tophits %>% 
  dplyr::filter(trait=="domestication") %>% 
  dplyr::pull(ME)

cor.turquoise <- datExpr[,geneInfo[,"moduleColor"]=="turquoise"]
cor.black <- datExpr[,geneInfo[,"moduleColor"]=="black"]
dim(cor.black)
```

```{r gene_sig_intramodular_connectivity}
datME=moduleEigengenes(datExpr,moduleColors)$eigengene

# module membership value for each gene in each module
datKME=signedKME(datExpr, datME, outputColumnName="MM.")
```

```{r}
hubGenes <- chooseTopHubInEachModule(datExpr = datExpr,
                         colorh = moduleColors, 
                         omitColors = "grey",
                         power = 2,
                         type = "signed"
                         )

hubGeneSignificance(datKME,
                    GSPvalue %>% dplyr::pull(p.GS.domestication))
# select top hub gene 
hubGenes[ME.tophits %>%
           dplyr::filter(trait=="domestication") %>% 
           dplyr::pull(ME) %>% gsub("ME","",.)] %>% as.data.frame()

```

```{r}
adjacency = adjacency(datExpr, power = softPower)
```



```{r}
#load("wgcna.RData")
save.image(file="wgcna.RData")
```

# SessionInfo
```{r}
sessionInfo()
```

